MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2022-02-12 main/MSstatsPTM_Main.tex
Updates to text

Generally, there are two main biological questions that are relevant in PTM analysis. First, some researchers run mass-spectrometry based experiments targeting PTMs when looking for biomarkers\cite{Petushkova_2017}. In this case the researcher may not necessarily want to correct for confounding with the total protein. The interest is in detecting modified peptides which change in abundance in a meaningful way, such as between disease and control, fast or slow-progressing, or responder and non-responder groups. Conversely, researchers may be interested in whether the fraction of peptides modified at a given site is changing between conditions. This is an important question for thinking about biological and physiological models and very short timescale experiments. In this case, we are interested in the relative abundance change solely from the modification and correcting for confounding with the total protein is relevant. These two questions are both important areas of study and require different statistical methods to correctly address. In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods as we must address confounding with the total protein.

@@

We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The spike-in peptides were mixed with human lysate to create four mixture conditions. Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only. The four mixtures were compared with known fold changes for the spike-in peptides.

@@

The proposed approach first leverages the large amount of work done to create MSstats~\cite{Choi:2014}. It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}. This approach allows the method to be extended to cases where the experimental design is unbalanced, and where additional sources of variation are present. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

Which amounts to running a the process without accounting for protein adjustment, as seen in MSstats and MSstatsTMT \cite{Choi:2014} \cite{Huang:2020}.

@@

The code is available on Bioconductor, http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html, and Github, https://github.com/Vitek-Lab/MSstatsPTM.

@@

The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. For the experiments where ground truth is known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites. These results were then compared to the ground truth to determine TP and TN. We then used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)). Finally, the summary statistics were compared across methods to analyze method performance.

For biological experiments where ground truth was not known, we again adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites. Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots. Each biological dataset presents a different experimental scenario, showing the applicability of the proposed method. All datasets in this paper were evaluated using the $MSstatsPTM$ package.

\subsubsection*{The proposed approach corrects for high FDR in computer simulations}

The proposed approach was evaluated against two-sample t-test and Limma. 
In Figure \ref{fig:fdr_all_models} we see that not accounting for changes in unmodified peptide level results in a high false discovery rate. When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR. Recall and accuracy of the proposed approach and Limma performed similarly, with Limma performing slightly better with low replicates, while t-test lagged behind. Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power with increased number of conditions.

\subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low replicates}

In this simulation, changes in unmodified peptide level still needed to be accounted for to control the FDR. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed strongest across all methods, as seen in Figure \ref{fig:sim_accuracy}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. Finally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. In comparison, Limma and t-test showed wider distributions. This fold change comparison can be seen in Figure \ref{fig:sim_boxplot}.

@@

%\paragraph*{Computer simulation: batch effects.}

%PTM experiments often create batches during data acquisition. The proposed approach can be naturally extended to handle such scenarios by modeling various forms of batch effects, as a t-test is not directly applicable to a problem with batches of data. Two commonly used ad-hoc approaches are a) t-test (no batch): ignoring batch effects when applying t-test, and b) t-test (most significant batch): applying t-test in each batch and drawing conclusions based on the most significant batch.\todo{Find source for batch analysis} Although simple, these ad-hoc methods lack statistical justification. We characterized their statistical properties under various forms of batch effects in Supplementary \todo{link to Sec. S4.2}. Two batches of data were generated, where various forms of batch effects were simulated, including difference in signal intensities across batches, difference in variability across batches, and interaction between batch and condition (i.e., change between conditions affected by batch).

%The major results are summarized in \todo{figref Fig. 5} with additional details provided in Supplementary \todo{link to Sec. S4.2}. In terms of estimation of fold change, the most statistically significant batch with t-test was highly variable and frequently biased (Fig. 5a). In the detection of systematic changes in PTM abundance, it also frequently resulted in higher false positive rates than the expected value (Fig. 5b). As shown Fig. 5c, the proposed approach improved power with small sample sizes in almost all the scenarios, under various forms of batch effects including difference in intensity level and difference in variability. Using t-test while ignoring batch effect gave similar performance in special cases with no difference in intensity level between batches (Supplementary Fig. \todo{find and link this section SXX}), but its performance dramatically decreased in general cases (Supplementary \todo{link to Figs. S8-S9}). The negative impact was only partially reduced by increasing the sample size to 5. The proposed approach gave consistently improved performance compared with other methods by properly characterizing batch effects and leveraging all available information.

@@

In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level. After adjustment the estimated fold change is more in line with expectation. Additionally, the background peptides serving as the null model, show many false positives before adjustment is made. After adjustment the results improve and the number of false positives decrease significantly.

In Figure \ref{fig:spikein_limma_volcano} we can see the results of the same experiment modeled using Limma with protein adjustment. Again the red labeled spike-in peptides are more inline with expected fold change after adjustment is applied. However, using Limma we can see that the majority of the differential spike in peptides are not showing a significant adjusted pvalue. In this case, using Limma would have resulted in missing the majority of differential modified peptides.  In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer afterward.

Further results of this experiment, including a comparison to t-test and fold change distribution of each method before and after adjustment, are provided in Supplementary Sec. \ref{supp-sec:benchmark}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection*{Dilution experiment [What was this experiment?]}

%We evaluated our approach using a custom designed dilution benchmark experiment where two sets of liver samples from five Atg16L1 deficient mice (one with KGG enrichment followed by trypsin digestion, one with trypsin digestion alone) were prepared. Following a randomized order, each sample was injected at three different concentrations 4ul, 2ul, and 1ul. The experiment resulted in 15 KGG and 15 global profiling runs. To correct for run-to-run variation, the AQUA peptide mixture was spiked into each sample. There are 4696 ubiquitinated proteins identified in this dataset. Among these, 3173 have corresponding measurements from global data.
%To simulate the effects of changes in protein abundance, we created three groups of data by matching one concentration for the KGG data with one for the global data: K1P1, K2P2, and K4P4. Systematic changes in the abundance of ubiquitinated sites between the groups are present in the KGG data. These changes are considered as artifacts driven by the changes in the abundance of their corresponding proteins. Using the comparison of K2P2 vs. K1P1 as example, there are 19355 possible comparisons with KGG data, and 3860 changes are detected. Out of the 19355 sites, 15036 have corresponding measurements from global data and are eligible for protein-level correction. The correction significantly reduced the number of false positives to 79. Similar observations were made in the other two comparisons K4P4 vs. K2P2 and K4P4 vs. K1P1, as shown in Table XX. \todo{Create or find this table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{MSstatsPTM corrects for bias in group comparison TMT experiment}

The results of this experiment are summarized in Figures \ref{fig:ipah_venn_diagramm} and \ref{fig:ipah_profile}. The experiment was modeled as a group comparison. In Figure \ref{fig:ipah_venn_diagramm} the number of significant modified peptides before and after adjustment is shown. We can see that more peptides become insignificant after adjustment than become significant. 3,236 peptides become insignificant, 1352 peptides become significant, while 4,282 peptides are significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues are correct in adjustment, and the true abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.

In Figure \ref{fig:ipah_profile}, looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method. The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This is contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e-8$. In this case the effect of the modified peptide is strongly confounded with changes in the global protein. The proposed method allows us to remove this confounding and estimate the true effect.

@@

The results of this experiment are summarized in Figures \ref{fig:shig_venn_d} and \ref{fig:shigella_profile}. The experiment was modeled as a time series, with the same bioreplicate measured at each time point. In Figure \ref{fig:shig_venn_d} the number of significant modified peptides before and after adjustment is shown. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides become insignificant, 4,947 became significant, while 41,552 peptides are significant in both models. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance. This is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

In Figure \ref{fig:shigella_profile} the profile plot of two specific modified peptides, $TTP\_MOUSE|P22893\_S178$ and $KI67\_MOUSE|E9PVX6\_T215$, show the power of the proposed method.  The profile plot of modified peptide $TTP\_MOUSE|P22893\_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions. However, the global profiling run shows a similar log fold change of $2.014$ between the same conditions. This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein. When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.248$. In contrast $KI67\_MOUSE|E9PVX6\_T215$ originally shows a small positive log fold change of $.185$ between conditions WT\_Early and WT\_Uninfected. This is contrast with a negative fold change in the global profiling run of $-.619$. In adjusting the modified peptide for changes in the global protein, the log fold change becomes $.804$ and the adjusted pvalue goes from $.338$ to $.0156$.

@@

As discussed previously, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a significant lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted. Additionally, not having a separate global profiling run resulted in very low feature counts for the unmodified protein model.

The results of this experiment are summarized in Figures \ref{fig:usp30_ven_d} and \ref{fig:usp30_profile}. In \ref{fig:usp30_ven_d} the number of significant modified peptides before and after protein adjustment is shown. In the top plot we can see there are a very large number of peptides that become insignificant after adjustment. This is mainly due to the lack of overlap between modified and unmodified peptides, stemming from not having a global profiling run. In the bottom plot only modified peptides that could be adjusted are shown. Here there are much less peptides that become insignificant after adjustment. In this plot we see that 726 peptides became insignificant, 547 became significant, and 1,078 are significant in both models. As in the previous experiments, we can check if these peptides are becoming insignificant due to an increase in variance. To check this we look at modified peptides whose adjusted fold change is within 10\% of their unadjusted fold change and they become insignificant after adjustment. In this experiment there are only 25 peptides that meet this criteria, a very small percentage of all modified peptides.

In Figure \ref{fig:usp30_profile} the profile plot of peptide $P52209\_K059$ is shown. The abundance between conditions appears flat in the modified peptide profile plot, however this is contrast with a large negative change in the corresponding global protein. Specifically looking at the comparison between the CCCP and Control conditions, the unadjusted model showed a small log fold change of $-.255$ and insignificant adjusted pvalue of $.104$. However when adjusting for changes in the undmodified protein, the log fold change increased to $-2.73$ and adjusted pvalue became significant at $1.067e-04$. Using the proposed method we get increased information that would otherwise have been missed.

@@

We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs. In terms of the number of replicates, we tested scenarios with equal replicates in both the modified and unmodified runs, as well as scenarios where the replicates differed between runs. For variance we used the biological experiments to determine what scenarios and values to test. In datasets 4 and 5 the variance of the PTM was higher than the global protein. In dataset 6 the variance of the PTM and Protein was generally the same due to not using a separate global profiling run. We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the protein and when they were equal. When the PTM and protein were the same we chose a variance of .15, whereas when the PTM was higher than the protein we chose a PTM variance of .2 and a protein variance of .1.

The results of the power and sample size analysis can be seen in Figure \ref{fig:power_sd_combo}. We can see that when the variance and replicates were equal, higher replicates predictably lead to higher power. In cases where the replicates were unbalanced, but the variance was still the same, it did not matter if there were more replicates in the modified or unmodified runs. As long as there were the same overall number of replicates the results were the same. In comparison, with differing variance and equal replicates, higher replicates still lead to higher power. When the replicates were unbalanced and the variance was higher in the PTM, there was more power when there were more replicates in the PTM than the protein. It is clearly more important to have high replicates for the PTM run than the unmodified protein, when the PTM variance is higher. In cases where the number of replicates has to be limited, it is better to weigh them on the PTM side.

@@

Additionally, expression levels of PTM sites can be convoluted with each other if there are two or more modification sites per peptide. In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified. One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution. However, this method would most likely run into major challenges due to sparsity of features for modified peptides with both a single and multiple modification sites. A more complex approach to addressing this problem is most likely necessary.

Overall, the proposed approach balances accuracy and practicality, and enables the analysis of complex experiments in high throughput. Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry). We are also interested in characterizing the interplay of PTMs at multiple sites.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.65]{images/fig3.png}
\caption{
Data structure of a typical PTM experiment and goals of PTM characterization. (a) Schematic data representation, in a simplified case of two conditions and two replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i*. Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). Some spectral features can be missing. (b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}

>>>>

There are two main biological questions that may be relevant in PTM analysis. First, some researchers run mass-spectrometry based experiments targeting PTMs when looking for biomarkers.\cite{Petushkova_2017}. In this case the researcher may not necessarily want to correct for confounding with the total protein. The interest is in detecting modified peptides which change in abundance in a meaningful way, such as between disease and control, fast or slow-progressing, or responder and non-responder groups. Conversely, researchers may be interested in whether the fraction of peptides modified at a given site is changing between conditions. This is an important question for thinking about biological and physiological models and very short timescale experiments. In this case, we are interested in the relative abundance change solely from the modification and correcting for confounding with the total protein is relevant. These two questions are both important areas of study and require different statistical methods to correctly address. In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods that account for confounding with the total protein.

@@

We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. Background E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The background lysate was treated as controls and are not expected to be differential in any comparison. The spike-in peptides were mixed with human lysate to create four mixture conditions. Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only. The four mixtures were compared with known fold changes for the spike-in peptides.

More information on the experimental design of the benchmark experiment can be found in Supplementary Sec. \ref{supp-sec:benchmark}.

@@

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}. This approach allows the method to be extended to cases where the experimental design is unbalanced, and where additional sources of variation are present. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

Which amounts to running the methods seen in MSstats and MSstatsTMT on peptide level data \cite{Choi:2014} \cite{Huang:2020}.

@@

The code is available on Bioconductor, \url{http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}, and Github, \url{https://github.com/Vitek-Lab/MSstatsPTM}.

@@

The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. For the experiments where ground truth is known, we calculated the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). True positives are modified peptides that are differential after accounting for changes in global protein abundance. True negatives were modified peptides that were not differentially abundant after accounting for global protein abundance. To determine positives and negatives we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites. These results were then compared to the ground truth to determine TP and TN. We used these results to calculate the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)). Finally, the summary statistics were compared across methods to analyze method performance.

For biological experiments where ground truth was not known, we adjusted the p-values for multiple testing and used an $\alpha=.05$ cutoff to determine differential sites. Comparisons were made before and after adjusting for changes in unmodified protein abundance and differences between the comparisons were noted. Each biological dataset presents a different experimental scenario, showing the applicability of the proposed method on a variety of experimental designs. All datasets in this paper were evaluated using the $MSstatsPTM$ package.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The proposed approach corrects for high FDR in computer simulations}

The proposed approach was evaluated against two-sample t-test and Limma in computer simulations. 
In Figure \ref{fig:computer_sim} we see that not accounting for changes in unmodified peptide level results in a high false discovery rate. When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR. In a clean simulation, recall and accuracy of the proposed approach and Limma performed similarly, with Limma performing slightly better with low replicates. In contrast, the performance of the t-test method lagged behind the others. Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data, leading to similar performance across all simulations. In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power in simulations with more conditions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low replicates}

In this simulation the proposed method outperformed t-test and limma when real world experimental conditions were introduced. Changes in unmodified peptide level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified peptide, the proposed method out performed the other methods, as seen in Figure \ref{fig:computer_sim}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. Additionally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change than t-test and Limma. Specifically, the inner quartile range of the estimated fold change for the proposed method was on average 10.4\% smaller than t-test and 21.8\% smaller than Limma. This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.

@@

@@

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance. After adjustment, the spike-in peptides' fold change was in line with expectation. This can be seen in Figure \ref{fig:spikein_boxplot}. While the median fold change for the spike-in peptides was generally in line with the true fold change after adjustment, we can see that the distribution of fold changes for the peptides was visibly wider after adjustment. However, the proposed method a tighter distribution around the true log fold change. Comparing the inner quartile range (IQR) of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than t-test's. This means that the proposed method's fold change estimate was more consistently correct, when compared to Limma and t-test.

In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein level. After adjustment the estimated fold change is more in line with expectation. Additionally, the background peptides serving as the null model, show many false positives before adjustment is made. After adjustment the results improve and the number of false positives decrease significantly. Specifically, for the proposed method, the number of false positives goes from 20.88\% to 1.84\% of all true negatives after adjustment is applied. Using both the proposed method and Limma, the red labeled spike-in peptides are more inline with expected fold change after the adjustment is applied. However, when using Limma we can see that the majority of the differential spike-in peptides are not showing a significant adjusted pvalue. In this case, using Limma would have resulted in missing the majority of differentially modified peptides.

Further results of this experiment are provided in Supplementary Sec. \ref{supp-sec:benchmark}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{MSstatsPTM corrects for bias in group comparison TMT experiment}

The results of this experiment are summarized in Figure \ref{fig:data4_plots}. The experiment was modeled as a group comparison. In Figure \ref{fig:data4_venn_diagram} the number of significant modified peptides before and after adjustment is shown. We can see that more peptides become insignificant after adjustment than become significant. 3,236 peptides become insignificant, 1352 peptides become significant, while 4,282 peptides are significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues are corrected in adjustment, and the true abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.

In Figure \ref{fig:data4_profile_plot}, looking at the modification of protein $GSDMD$ at $K62$ we can see the advantage of the proposed method. The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This is contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became significant, $5.25e-8$. In this case the effect of the modified peptide is strongly confounded with changes in the global protein. The proposed method allows us to remove this confounding and estimate the true effect.

@@

The results of this experiment are summarized in Figure \ref{fig:data5_plots}. The experiment was modeled as a time series, with the same biological replicate measured at each time point. Modeling the experiment as a time series lowered the model variance as compared to modeling the experiment as a group comparison. In Figure \ref{fig:data5_venn_diagram} the number of significant modified peptides before and after adjustment is shown. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides become insignificant, 4,947 became significant, while 41,552 peptides are significant in both models. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance. This is still a small portion (3.4\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

In Figure \ref{fig:data5_profile_plot} the profile plot of protein $TTP$ modified at site $S178$ shows the power of the proposed method.  The profile plot shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions. However, the global profiling run shows a similar log fold change of $2.014$ between the same conditions. This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein. When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.248$. Correcting for changes in global protein abundance allowed us to see the true impact of the modification at $S178$ which would have otherwise been challenging to perceive.

Analysis of other modifications in this experiment can be seen in Supplementary Sec. \ref{supp-sec:shigella}.

@@

As discussed previously, there was no unmodified global profiling run performed in this experiment. Once identification and quantification of the Ubiquitination profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a significant lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified protein and could be adjusted. Additionally, not having a separate global profiling run resulted in low feature counts for the unmodified protein model.

The results of this experiment are summarized in Figure \ref{fig:data6_plots}. In \ref{fig:data6_vd1} the number of significant modified peptides before and after protein adjustment is shown. In the top plot we can see there are a large number of peptides that become insignificant after adjustment. This is mainly due to the lack of overlap between modified and unmodified peptides, stemming from not having a global profiling run. In the bottom plot only modified peptides that could be adjusted are shown. Here there are much less peptides that become insignificant after adjustment. In this plot we see that 726 peptides became insignificant, 547 became significant, and 1,078 are significant in both models. As in the previous experiments, we can check if these peptides are becoming insignificant due to an increase in variance. To check this we look at modified peptides whose adjusted fold change is within 10\% of their unadjusted fold change and they become insignificant after adjustment. In this experiment there are only 25 peptides that meet this criteria, a small percentage of all modified peptides.

Analysis of modifications before and after protein adjustment can be seen in Supplementary Sec. \ref{supp-sec:usp30}.

%In Figure \ref{fig:usp30_profile} the profile plot of peptide $P52209\_K059$ is shown. The abundance between conditions appears flat in the modified peptide profile plot, however this is contrast with a large negative change in the corresponding global protein. Specifically looking at the comparison between the CCCP and Control conditions, the unadjusted model showed a small log fold change of $-.255$ and insignificant adjusted pvalue of $.104$. However when adjusting for changes in the undmodified protein, the log fold change increased to $-2.73$ and adjusted pvalue became significant at $1.067e-04$. Using the proposed method we get increased information that would otherwise have been missed.

@@

We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs. In terms of the number of replicates, we tested scenarios with equal replicates in both the modified and unmodified runs, as well as scenarios where the replicates differed between runs. For variance we used the biological experiments to determine what scenarios and values to test. In datasets 4 and 5 the variance of the PTM was higher than the global protein. In dataset 6 the variance of the PTM and Protein was generally the same due to not using a separate global profiling run. We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the protein and when they were equal. When the PTM and protein were the same we chose a variance of $.15$, whereas when the PTM was higher than the protein we chose a PTM variance of $.2$ and a protein variance of $.1$.

The results of the power and sample size analysis can be seen in Supplementary Figure \ref{supp-fig:power_sd_combo}. When the variance and replicates were equal, higher replicates predictably lead to higher power. In cases where the replicates were unbalanced, but the variance was still the same, it did not matter if there were more replicates in the modified or unmodified runs. As long as there were the same overall number of replicates the results were the same. In comparison, with differing variance and equal replicates, higher replicates still lead to higher power. When the replicates were unbalanced and the variance was higher in the PTM, there was more power when there were more replicates in the PTM than the protein. It is clearly more important to have high replicates for the PTM run than the unmodified protein, when the PTM variance is higher. In cases where the number of replicates has to be limited, it is better to weigh them on the PTM side.

@@

Additionally, expression levels of PTM sites can be convoluted with each other if there are two or more modification sites per peptide. In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified. One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution. However, this method would likely run into challenges due to sparsity of features for modified peptides with both a single and multiple modification sites. A more complex approach to addressing this problem is most likely necessary.

Overall, the proposed approach balances accuracy and practicality, and enables the analysis of complex experiments in high throughput. Future work is to carry out the inference and testing for not only the relative change of PTM abundance, but also the fraction of the protein that is modified at the particular site (site occupancy, or stoichiometry), and attempt to deconvolute the effect of individual PTMs in peptides with multiple modifications.

\section{Data Availability}

Shigella: https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.6]{images/fig3.png}
\caption{
Data structure of a typical PTM experiment and goals of PTM characterization. (a) Schematic data representation, in a simplified case of two conditions and two replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i*. Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). Some spectral features can be missing. (b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}
